---
title: "Pacific Reef Egret - prelim data"
author: "NLJ - MvdP - GSC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE}
library(readxl)
library(openxlsx)
library(tidyverse)
library(viridis)
library(raster)
library(amt)
library(purrr)
library(adehabitatHR)
library(Polychrome)
library(lubridate)
library(dataaimsr)
library(sf)
library(janitor)
library(plyr)
library(skimr)
library(glmmTMB)
library(effects)
library(kableExtra)



# shp

gbrFeat<-st_read("./data/GBR_FEATURES/egFeats.shp", quiet=TRUE)%>%
  st_transform(4326)

```

There are some points as far as Curtis island (n=\~10) from several individuals but I have concentrated on those that will tell us about the foraging behaviour of PRE. If we had more dispersal data that mught be an interesting investigation but we really have very few points outside the Wistari-Heron-One Tree area. Incredibly high foraging site fidelity for all birds for the most part.

GPS fixes per band number:

(2190 fixes = \~1 year of data at 4 hour fixes)

```{r 1&2trackData, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width='100%'}



source("./src/1.egTracks_HI_Wist.R")
source("./src/2.egTracks_OTI.R")


preDat <- rbind(pre, oti_pre) %>%
  dplyr::mutate(BandNumber = as.character(BandNumber)) %>%
  dplyr::filter(
    x > 151.6, x < 152.3,
    y > -23.7, y < -23.3
  ) %>%
  dplyr::distinct(BandNumber, dateTimeAEST, .keep_all = TRUE) %>%
  droplevels() %>%
  dplyr::arrange(BandNumber, dateTimeUTC) %>%
  dplyr::group_by(BandNumber) %>%
  dplyr::mutate(
    tDiff = difftime(lead(dateTimeUTC), dateTimeUTC, units = "hours"),
    lagdateTimeUTC = lag(dateTimeUTC)
  )


#skimr::skim(preDat)

preDat %>%
  group_by(BandNumber)%>%
  dplyr::summarise(fixes_n = n(),
                   Location = first(LOCBANDED),
                   Sex = first(SEX),
                   Morph = first(MORPH),
                   MinTimeGap = min(tDiff, na.rm=TRUE),
                   MeanTimeGap = mean(tDiff, na.rm=TRUE),
                   MaxTimeGap = max(tDiff, na.rm=TRUE))%>%
  kable()

saveRDS(preDat,"./src_outputs/cleanTracks.rds")

ggplot()+
  geom_sf(data=gbrFeat)+
  geom_point(data=preDat, aes(x=x, y=y), size=0.8, col='red')+
  facet_wrap(~BandNumber, ncol = 5)+
  ylim(c(-23.7,-23.17))+
  xlim(c(151.6,152.2))

```

Bimodal elevation distribution likely indicating reef and island positions

```{r 3DEMdat, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width='50%'}


source("./src/3.demData.R")

ggplot() +
  geom_histogram(data = extrDEM, aes(x = finalElevation), bins = 60, fill = "steelblue", color = "black") +
  labs(x = "Elevation (m)", y = "Count", title = "Distribution of Elevation") +
  theme_minimal()+
  xlim(c(-10,10))

extrDEM%>%
  dplyr::filter(finalElevation>-5)%>%
ggplot(aes(x = "", y = finalElevation)) +  # "" to avoid categorical x-axis
  geom_violin(fill = "lightblue", color = "black") +  # Violin plot style
  labs(y = "Elevation (m)", title = "Elevation Distribution") +
  theme_minimal()

saveRDS(extrDEM, "./src_outputs/finalDepth.rds") 


```

Deeper than expected foraging depth however, this may be a case of Egrets using small features such as bommies on which to forage that are set in otherwise deeper areas.

Distribution of elevation used and the influence of tide and time of day. Some indication that all birds prefer to remain well above the low tide line. No sex-size dimorphism obvious nor influences of feather colour (morph) on prefered elevation under different tide regimes (spring/neap)

```{r 4tideDat, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width='50%'}

rm(list=ls())

source("./src/4.tides.R")


# might almost make sense to look at it from the below ASL vs not
# as a logistic reg.?

tidepts%>%
  dplyr::filter(finalElevation>-5 & finalElevation<1.2)%>%
ggplot() +
  geom_point(aes(x=nearestTide, y=finalElevation, col=SEX)) +
  geom_smooth(aes(x=nearestTide, y=finalElevation), col='blue')+
  theme_minimal()+
  labs(x="Tide Height", y="Elevation")

tidepts%>%
  dplyr::filter(finalElevation>-5 & finalElevation<1.2)%>%
ggplot() +
  geom_point(aes(x=nearestTide, y=finalElevation, col=MORPH)) +
  geom_smooth(aes(x=nearestTide, y=finalElevation), col='blue')+
  theme_minimal()+
  labs(x="Tide Height", y="Elevation")

tidepts%>%
  dplyr::filter(finalElevation>-5 & finalElevation<1.2)%>%
ggplot() +
  geom_point(aes(x=nearestTide, y=finalElevation, col=LOCBANDED)) +
  geom_smooth(aes(x=nearestTide, y=finalElevation), col='blue')+
  theme_minimal()+
  labs(x="Tide Height", y="Elevation")


tidepts %>%
  dplyr::filter(actualDepth > -5) %>%  # Filter depths greater than -5m
  ggplot(aes(x = "", y = actualDepth)) +  
  geom_violin(fill = "lightblue", color = "black", adjust = 1) +  # Violin plot style
  scale_y_continuous(
    breaks = seq(floor(min(tidepts$actualDepth, na.rm = TRUE)), 
                 ceiling(max(tidepts$actualDepth, na.rm = TRUE)), 
                 by = 1)  # Label every integer
  ) +
  labs(y = "Actual Depth (m)", 
       title = "Distribution of Depths Selected", 
       subtitle = "Actual depth = Tide - Elevation") +
  theme_minimal()



ggplot()+
  geom_point(data=tidepts, aes(x= moonPhase, y= nearestTide, col=nearestTideCat), alpha=0.3)+
  scale_colour_manual(values=c(wesanderson::wes_palette("Darjeeling1",5,'discrete')[c(1,3,5)]))+
  facet_grid(~neapSpring)+
  theme_minimal()



tidepts%>%
  dplyr::filter(actualDepth<2 & actualDepth>-2)%>%
  ggplot()+
  geom_boxplot(aes(x=TOD, y=actualDepth, group=TOD))+
  geom_smooth(aes(x=TOD, y=actualDepth))+
  theme_minimal()+
  labs(y = "Elevation (m)")

tidepts%>%
  dplyr::filter(finalElevation<2 & finalElevation>-2)%>%
  ggplot()+
  geom_boxplot(aes(x=SEX, y=finalElevation))+
  facet_grid(nearestTideCat~neapSpring)+
  theme_minimal()+
  labs(y = "Elevation (m)")

tidepts%>%
  dplyr::filter(finalElevation<2 & finalElevation>-2)%>%
  ggplot()+
  geom_hline(yintercept =0, linetype="dotted")+
  geom_boxplot(aes(x=MORPH, y=finalElevation))+
  facet_grid(nearestTideCat~neapSpring)+
  theme_minimal()+
  labs(y = "Elevation (m)")


# to do:
# does foraging KDE change with moonphase i.e. greater area to forage?
# is time of first/last non-island point changed by moon fraction?





saveRDS(tidepts,"./src_outputs/tidesFinal.rds")

```

Indicates that there is a preference to be at higher elevations i.e. the islands during the dark hours. Target foraging depth appears to be just over 1m which is deeper than expected given the size of these birds.

```{r 5lightDark, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width='500%'}

rm(list=ls())

source("./src/5.LightDark.R")

lightPRE %>%
  dplyr::filter(actualDepth > -5) %>%  # Filter depths greater than -5m
  ggplot(aes(x = lightPeriod, y = actualDepth)) +  
  geom_violin(aes(fill = lightPeriod), color = "black", adjust = 1) +  # Violin plot style
  scale_y_continuous(
    breaks = seq(floor(min(lightPRE$actualDepth, na.rm = TRUE)), 
                 ceiling(max(lightPRE$actualDepth, na.rm = TRUE)), 
                 by = 1)  # Label every integer
  ) +
  scale_fill_manual(values=c("grey","gold"))+
  labs(y = "Actual Depth (m)", 
       title = "Distribution of Depths Selected", 
       subtitle = "Actual depth = Tide - Elevation") +
  theme_minimal()

lightPRE %>%
  dplyr::filter(actualDepth > -5) %>%  # Filter depths greater than -5m
  ggplot(aes(x = lightPeriod, y = finalElevation)) +  
  geom_violin(aes(fill = lightPeriod), color = "black", adjust = 1) +  # Violin plot style
  scale_y_continuous(
    breaks = seq(floor(min(lightPRE$finalElevation, na.rm = TRUE)), 
                 ceiling(max(lightPRE$finalElevation, na.rm = TRUE)), 
                 by = 1)  # Label every integer
  ) +
  scale_fill_manual(values=c("grey","gold"))+
  labs(y = "Elevation (m)", 
       title = "Distribution of Elevation Selected") +
  theme_minimal()


saveRDS(lightPRE,"./src_outputs/lightFinal.rds")




```

Raw counts of geomorphic and benthic data classes used from all points, island indicate locations on Heron or One Tree islands.

```{r 6benthosData, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE}


source("./src/6.BenthosType.R")

geodat%>%
ggplot()+
  geom_sf(data=geo, aes(fill=class))+
  geom_sf(data=lightPoints)+
  theme_minimal()

geodat%>%
ggplot()+
  geom_sf(data=benth, aes(fill=class))+
  geom_sf(data=lightPoints)+
  theme_minimal()

geodat%>%
  group_by(geomorphicCover, benthicCover,  MORPH)%>%
  dplyr::summarise(count=as.numeric(n()))%>%
  filter(!is.na(geomorphicCover), !is.na(benthicCover))%>%
  ungroup%>%
  ggplot(data=., aes(fill=benthicCover, y=count, x=geomorphicCover))+
  geom_bar(position="stack", stat="identity")+
  facet_grid(~MORPH)+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


geodat%>%
  group_by(geomorphicCover, benthicCover,  SEX)%>%
  dplyr::summarise(count=as.numeric(n()))%>%
  filter(!is.na(geomorphicCover), !is.na(benthicCover))%>%
  ungroup%>%
  ggplot(data=., aes(fill=benthicCover, y=count, x=geomorphicCover))+
  geom_bar(position="stack", stat="identity")+
  facet_grid(~SEX)+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


saveRDS(geodat, "./data/DFs/benthosFinal.rds")

```

Water temperatures as measured from data loggers at Heron Island

```{r 7waterTemp, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE}

source("./src/7.waterTemps.R")

all_joined%>%
ggplot()+
  geom_point(aes(x=month, y=flatTemp), col='brown', alpha=0.2) +
  geom_point(aes(x=month, y=slopeTemp), col='grey', aplha=0.2)+
  theme_minimal()+
  labs(x='Month', y="Temp DegC", title = "Reef Flat and Slope temperatures over multiple years - Heron Island",
       subtitle = "Brown = Reef Flat (1m), Grey = Reef Slope (8.5m)")



saveRDS(all_joined, "./src_outputs/finalTemp.rds")



```

Plotting AKDE (Auto-correlated KDE) we see an expected cyclic increase and decrease from one individual (101-18510) but few notable patterns from others, potentially due to the limited multi-year data sets. Potentially a hint of cycling between larger (winter) and smaller (summer) KDEs. Issues of why we have such data gaps remain unclear.

```{r kde, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, fig.height=16}

load('./src_outputs/allAKDE_monthly.Rdata')
load('./src_outputs/ctmmTelemetryMonthly.Rdata')

monthIDs<-names(tel)

areaRows <- lapply(allAKDE, function(x) {
  # Check if CI summary exists and if not, return a row of "NULL"
  if (inherits(x, "UD")) {  # Check if x is a valid 'UD' object
    tryCatch({
      ci_values <- summary(x)$CI
      # Return a row with the CI values: low, est, and high
      return(c(rownames(ci_values), ci_values[1, "low"], ci_values[1, "est"], ci_values[1, "high"]))
    }, error = function(e) {
      # If there's an error, return "NULL" values
      return(c("NULL", "NULL", "NULL", "NULL"))
    })
  } else {
    # If x is not a valid 'UD' object, return "NULL"
    return(c("NULL", "NULL", "NULL", "NULL"))
  }
})


areaDF <- do.call(rbind, areaRows)
colnames(areaDF) <- c("units", "CI_Lower", "CI_Estimate", "CI_Upper")
akdeDF <- cbind(monthIDs, areaDF) %>%
  data.frame() %>%
  tidyr::separate_wider_delim(
    cols = monthIDs,
    delim = "_",
    names = c("BandNumber", "date")
  ) %>%
  dplyr::mutate(
    date = lubridate::ymd(date),
    CI_Lower = as.numeric(CI_Lower),
    CI_Estimate = as.numeric(CI_Estimate),
    CI_Upper = as.numeric(CI_Upper),
    
    # Convert units to square kilometers
    CI_Lower = dplyr::case_when(
      grepl("square centimeters", units) ~ CI_Lower / 1e10,
      grepl("square meters", units) ~ CI_Lower / 1e6,
      grepl("hectares", units) ~ CI_Lower / 100,
      TRUE ~ CI_Lower
    ),
    CI_Estimate = dplyr::case_when(
      grepl("square centimeters", units) ~ CI_Estimate / 1e10,
      grepl("square meters", units) ~ CI_Estimate / 1e6,
      grepl("hectares", units) ~ CI_Estimate / 100,
      TRUE ~ CI_Estimate
    ),
    CI_Upper = dplyr::case_when(
      grepl("square centimeters", units) ~ CI_Upper / 1e10,
      grepl("square meters", units) ~ CI_Upper / 1e6,
      grepl("hectares", units) ~ CI_Upper / 100,
      TRUE ~ CI_Upper
    ),
    
    # Update units column
    units = dplyr::case_when(
      grepl("square centimeters", units) ~ "area (square kilometers)",
      grepl("square meters", units) ~ "area (square kilometers)",
      grepl("hectares", units) ~ "area (square kilometers)",
      TRUE ~ units
    )
  ) %>%
  dplyr::arrange(BandNumber, date)




ggplot()+
  geom_point(data=akdeDF, aes(x=date, y=CI_Estimate))+
  geom_line(data=akdeDF, aes(x=date, y=CI_Estimate))+
  facet_wrap(~BandNumber, ncol = 1)+
  labs(x='Month',y="95% KDE area estimates (km^2)",
       title = "AKDE estimates per month, per individual")+
  theme_bw()










```
