---
title: "Pacific Reef Egret - prelim data"
author: "NLJ - MvdP - GSC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE}
library(readxl)
library(openxlsx)
library(tidyverse)
library(viridis)
library(raster)
library(amt)
library(purrr)
library(adehabitatHR)
library(Polychrome)
library(lubridate)
library(dataaimsr)
library(sf)
library(janitor)
library(plyr)
library(skimr)
library(glmmTMB)
library(effects)
library(kableExtra)



# shp

gbrFeat<-st_read("./data/GBR_FEATURES/egFeats.shp", quiet=TRUE)%>%
  st_transform(4326)

```

There are some points as far as Curtis island (n=\~10) from several individuals but I have concentrated on those that will tell us about the foraging behaviour of PRE. If we had more dispersal data that mught be an interesting investigation but we really have very few points outside the Wistari-Heron-One Tree area. Incredibly high foraging site fidelity for all birds for the most part.

GPS fixes per band number:

(2190 fixes = \~1 year of data at 4 hour fixes)

```{r 1&2trackData, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width='100%', fig.height=c(15,10)}


source("./src/1.egTracks_HI_Wist.R")
source("./src/2.egTracks_OTI.R")


preDat <- rbind(pre, oti_pre) %>%
  dplyr::mutate(BandNumber = as.character(BandNumber)) %>%
  dplyr::filter(
    x > 151.6, x < 152.3,
    y > -23.7, y < -23.3
  ) %>%
  dplyr::distinct(BandNumber, dateTimeAEST, .keep_all = TRUE) %>%
  droplevels() %>%
  dplyr::arrange(BandNumber, dateTimeUTC) %>%
  dplyr::group_by(BandNumber) %>%
  dplyr::mutate(
    tDiff = difftime(lead(dateTimeUTC), dateTimeUTC, units = "hours"),
    lagdateTimeUTC = lag(dateTimeUTC)
  )

preDat %>%
  group_by(BandNumber)%>%
  dplyr::summarise(fixes_n = n(),
                   Location = first(LOCBANDED),
                   Sex = first(SEX),
                   Morph = first(MORPH),
                   MinTimeGap = min(tDiff, na.rm=TRUE),
                   MeanTimeGap = mean(tDiff, na.rm=TRUE),
                   MaxTimeGap = max(tDiff, na.rm=TRUE))%>%
  kable()

saveRDS(preDat,"./src_outputs/cleanTracks.rds")

preDat%>%
  mutate(week = lubridate::floor_date(dateTimeUTC,'week'))%>%
  group_by(BandNumber, week)%>%
  dplyr::summarise(weekPoints = n())%>%
  ungroup%>%
ggplot(., aes(x=week, y=weekPoints))+
  geom_point()+
  facet_wrap(~BandNumber, ncol = 1)

chk<-preDat%>%
  mutate(date = date(dateTimeUTC))%>%
  group_by(BandNumber, date)%>%
  dplyr::summarise(dayPoints = n())%>%
  ungroup

ggplot()+
  geom_sf(data=gbrFeat)+
  geom_point(data=preDat, aes(x=x, y=y), size=0.8, col='red')+
  facet_wrap(~BandNumber, ncol = 5)+
  ylim(c(-23.7,-23.17))+
  xlim(c(151.6,152.2))

```

Bimodal elevation distribution likely indicating reef and island positions

```{r 3DEMdat, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width='50%'}


source("./src/3.demData.R")

ggplot() +
  geom_histogram(data = extrDEM, aes(x = finalElevation), bins = 60, fill = "steelblue", color = "black") +
  labs(x = "Elevation (m)", y = "Count", title = "Distribution of Elevation") +
  theme_minimal()+
  xlim(c(-10,10))

extrDEM%>%
  dplyr::filter(finalElevation>-5)%>%
ggplot(aes(x = "", y = finalElevation)) +  # "" to avoid categorical x-axis
  geom_violin(fill = "lightblue", color = "black") +  # Violin plot style
  labs(y = "Elevation (m)", title = "Elevation Distribution") +
  theme_minimal()

saveRDS(extrDEM, "./src_outputs/finalDepth.rds") 


```

Deeper than expected foraging depth however, this may be a case of Egrets using small features such as bommies on which to forage that are set in otherwise deeper areas.

Distribution of elevation used and the influence of tide and time of day. Some indication that all birds prefer to remain well above the low tide line. No sex-size dimorphism obvious nor influences of feather colour (morph) on prefered elevation under different tide regimes (spring/neap)

```{r 4tideDat, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width='50%'}

rm(list=ls())

source("./src/4.tides.R")


# might almost make sense to look at it from the below ASL vs not
# as a logistic reg.?

tidepts%>%
  dplyr::filter(finalElevation>-5 & finalElevation<1.2)%>%
ggplot() +
  geom_point(aes(x=nearestTide, y=finalElevation, col=SEX)) +
  geom_smooth(aes(x=nearestTide, y=finalElevation), col='blue')+
  theme_minimal()+
  labs(x="Tide Height", y="Elevation")

tidepts%>%
  dplyr::filter(finalElevation>-5 & finalElevation<1.2)%>%
ggplot() +
  geom_point(aes(x=nearestTide, y=finalElevation, col=MORPH)) +
  geom_smooth(aes(x=nearestTide, y=finalElevation), col='blue')+
  theme_minimal()+
  labs(x="Tide Height", y="Elevation")

tidepts%>%
  dplyr::filter(finalElevation>-5 & finalElevation<1.2)%>%
ggplot() +
  geom_point(aes(x=nearestTide, y=finalElevation, col=LOCBANDED)) +
  geom_smooth(aes(x=nearestTide, y=finalElevation), col='blue')+
  theme_minimal()+
  labs(x="Tide Height", y="Elevation")


tidepts %>%
  dplyr::filter(actualDepth > -5) %>%  # Filter depths greater than -5m
  ggplot(aes(x = "", y = actualDepth)) +  
  geom_violin(fill = "lightblue", color = "black", adjust = 1) +  # Violin plot style
  scale_y_continuous(
    breaks = seq(floor(min(tidepts$actualDepth, na.rm = TRUE)), 
                 ceiling(max(tidepts$actualDepth, na.rm = TRUE)), 
                 by = 1)  # Label every integer
  ) +
  labs(y = "Actual Depth (m)", 
       title = "Distribution of Depths Selected", 
       subtitle = "Actual depth = Tide - Elevation") +
  theme_minimal()



ggplot()+
  geom_point(data=tidepts, aes(x= moonPhase, y= nearestTide, col=nearestTideCat), alpha=0.3)+
  scale_colour_manual(values=c(wesanderson::wes_palette("Darjeeling1",5,'discrete')[c(1,3,5)]))+
  facet_grid(~neapSpring)+
  theme_minimal()



tidepts%>%
  dplyr::filter(actualDepth<2 & actualDepth>-2)%>%
  ggplot()+
  geom_boxplot(aes(x=TOD, y=actualDepth, group=TOD))+
  geom_smooth(aes(x=TOD, y=actualDepth))+
  theme_minimal()+
  labs(y = "Elevation (m)")

tidepts%>%
  dplyr::filter(finalElevation<2 & finalElevation>-2)%>%
  ggplot()+
  geom_boxplot(aes(x=SEX, y=finalElevation))+
  facet_grid(nearestTideCat~neapSpring)+
  theme_minimal()+
  labs(y = "Elevation (m)")

tidepts%>%
  dplyr::filter(finalElevation<2 & finalElevation>-2)%>%
  ggplot()+
  geom_hline(yintercept =0, linetype="dotted")+
  geom_boxplot(aes(x=MORPH, y=finalElevation))+
  facet_grid(nearestTideCat~neapSpring)+
  theme_minimal()+
  labs(y = "Elevation (m)")


# to do:
# does foraging KDE change with moonphase i.e. greater area to forage?
# is time of first/last non-island point changed by moon fraction?





saveRDS(tidepts,"./src_outputs/tidesFinal.rds")

```

Indicates that there is a preference to be at higher elevations i.e. the islands during the dark hours. Target foraging depth appears to be just over 1m which is deeper than expected given the size of these birds.

```{r 5lightDark, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width='50%'}

rm(list=ls())

source("./src/5.LightDark.R")

lightPRE %>%
  dplyr::filter(actualDepth > -5) %>%  # Filter depths greater than -5m
  ggplot(aes(x = lightPeriod, y = actualDepth)) +  
  geom_violin(aes(fill = lightPeriod), color = "black", adjust = 1) +  # Violin plot style
  scale_y_continuous(
    breaks = seq(floor(min(lightPRE$actualDepth, na.rm = TRUE)), 
                 ceiling(max(lightPRE$actualDepth, na.rm = TRUE)), 
                 by = 1)  # Label every integer
  ) +
  scale_fill_manual(values=c("grey","gold"))+
  labs(y = "Actual Depth (m)", 
       title = "Distribution of Depths Selected", 
       subtitle = "Actual depth = Tide - Elevation") +
  theme_minimal()

lightPRE %>%
  dplyr::filter(actualDepth > -5) %>%  # Filter depths greater than -5m
  ggplot(aes(x = lightPeriod, y = finalElevation)) +  
  geom_violin(aes(fill = lightPeriod), color = "black", adjust = 1) +  # Violin plot style
  scale_y_continuous(
    breaks = seq(floor(min(lightPRE$finalElevation, na.rm = TRUE)), 
                 ceiling(max(lightPRE$finalElevation, na.rm = TRUE)), 
                 by = 1)  # Label every integer
  ) +
  scale_fill_manual(values=c("grey","gold"))+
  labs(y = "Elevation (m)", 
       title = "Distribution of Elevation Selected") +
  theme_minimal()


saveRDS(lightPRE,"./src_outputs/lightFinal.rds")




```

Raw counts of geomorphic and benthic data classes used from all points, island indicate locations on Heron or One Tree islands.

```{r 6benthosData, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, out.width="50%"}


source("./src/6.BenthosType.R")

# Define color palette
my_cols <- wesanderson::wes_palette("Zissou1", n = length(unique(geodat$benthicCover)), type = "continuous")

# First plot: geomorphic classes
geodat %>%
  ggplot() +
  geom_sf(data = geo, aes(fill = class), col=NA) +
  geom_sf(data = lightPoints, shape = 21, fill = NA, color = "black", size = 1) +
  theme_minimal()

# Second plot: benthic classes
geodat %>%
  ggplot() +
  geom_sf(data = benth, aes(fill = class)) +
  geom_sf(data = lightPoints, shape = 21, fill = NA, color = "black", size = 1) +
  theme_minimal()

# Third plot: stacked bar by MORPH
geodat %>%
  group_by(geomorphicCover, benthicCover, MORPH) %>%
  dplyr::summarise(count = as.numeric(n())) %>%
  filter(!is.na(geomorphicCover), !is.na(benthicCover)) %>%
  ungroup() %>%
  ggplot(aes(fill = benthicCover, y = count, x = geomorphicCover)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_grid(~MORPH) +
  theme_minimal() +
  scale_fill_manual(values = my_cols) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

# Fourth plot: stacked bar by SEX
geodat %>%
  group_by(geomorphicCover, benthicCover, SEX) %>%
  dplyr::summarise(count = as.numeric(n())) %>%
  filter(!is.na(geomorphicCover), !is.na(benthicCover)) %>%
  ungroup() %>%
  ggplot(aes(fill = benthicCover, y = count, x = geomorphicCover)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_grid(~SEX) +
  theme_minimal() +
  scale_fill_manual(values = my_cols) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


saveRDS(geodat, "../../data/DFs/benthosFinal.rds")

```

Water temperatures as measured from data loggers at Heron Island

```{r 7waterTemp, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, fig.height=c(15,7)}

source("./src/7.waterTemps.R")

all_joined%>%
ggplot()+
  geom_point(aes(x=month, y=flatTemp), col='brown', alpha=0.2) +
  geom_point(aes(x=month, y=slopeTemp), col='grey', aplha=0.2)+
  theme_minimal()+
  labs(x='Month', y="Temp DegC", title = "Reef Flat and Slope temperatures over multiple years - Heron Island",
       subtitle = "Brown = Reef Flat (1m), Grey = Reef Slope (8.5m)")



saveRDS(all_joined, "./src_outputs/finalTemp.rds")



```

Plotting AKDE (Auto-correlated KDE) we see an expected cyclic increase and decrease from one individual (101-18510) but few notable patterns from others, potentially due to the limited multi-year data sets. Potentially a hint of cycling between larger (winter) and smaller (summer) KDEs. Issues of why we have such data gaps remain unclear.

```{r kde, message=FALSE, warning=FALSE, error=FALSE, echo=FALSE, fig.height=16}

akdeDFweekly<-readRDS("./src_outputs/akde_weeklyFinal.rds")

ggplot()+
  geom_point(data=akdeDFweekly, aes(x=date, y=CI_Estimate))+
  facet_wrap(~BandNumber, ncol = 1)

akdeDFmonthly<-readRDS("./src_outputs/akde_monthlyFinal.rds")


ggplot()+
  geom_point(data=akdeDF, aes(x=date, y=CI_Estimate))+
  facet_wrap(~BandNumber, ncol = 1)








```

```{r}

# Get 
# Departure time (first off-island fix)
# 
# Return time (last on-island fix)
# 
# Trip duration (time between first off-island and return)
# 
# Trip frequency (number of off-island trips per day)
# 
# Max distance from island
# 
# Total distance traveled

# 3. Breeding Signal: Temporal Pattern Changes
# 
# Nesting birds tend to:
# 
#     Leave later in the morning or take fewer trips per day
# 
#     Have shorter, more consistent foraging trips
# 
#     Travel less far (because one parent often stays near the nest or they alternate care)
# 
#     Spend more time stationary on the island, particularly overnight
# 
# You can:
# 
#     Plot daily foraging metrics over time and look for shifts or step changes
# 
#     Apply change-point analysis or a hidden Markov model to detect behavioral states
# 
#     Compare known pre-breeding and suspected breeding periods

# 4. Consistency at a Specific Point
# 
# If GPS accuracy allows (~10–20 m), and birds show high site fidelity to a specific spot on the island (same fix over multiple days), that can be a strong nesting signal.
# 
#     Filter for days with >3 fixes at the same coordinate or within ~10 m radius for >5 consecutive days.

# 5. Combine Movement and Timing Cues
# 
# If you have even rough colony phenology (e.g., when nesting usually starts), use that as a temporal filter. Then:
# 
#     Check if individuals start showing reduced movement + high site fidelity at that time.
# 
#     Cross-reference with diel activity patterns (less movement at night, more time on island, etc.).
    
```
